---
title: "R markdown"
author: "Hevar"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---
### R Markdown

## First Set of questions 
Reading the date and making adjustments 
```{r}
df <- read.csv('activity.csv')
df$date <- as.Date(df$date)
str(df)
colSums(is.na(df))
```

Calculate the total number of steps taken per day
```{r}
step_sum <- tapply(df$steps, df$date, sum, na.rm = T)
sum_data <- as.data.frame(step_sum)
sum_data$dates <- rownames(sum_data)
head(sum_data)

```

Creating the histogram
```{r}
colors <- colorRampPalette(c('red', 'blue'))
color <- colors(5)
hist(sum_data$step_sum,col = color, xlab = 'Sum of the steps', 
        main = 'Total Steps per Day')
rug(sum_data$step_sum, lwd = 2, lty = 1, col = 'orange')
```

Calculating the mean and median for the total steps per day
```{r pressure}
mean_value <- mean(sum_data$step_sum, na.rm = T)
median_value <- median(sum_data$step_sum, na.rm = T)
print(paste('The mean value of the total steps is :' ,round(mean_value, 2)))
print(paste('The median value of the total steps is :' ,median_value))
```

## Second Set Of Questions 

What is the average daily activity pattern?
```{r}
second_df <- df # this was done to avoid messing with the original data
interval_avg_steps <- aggregate(steps ~ interval, data = second_df, 
        FUN = mean, na.rm = TRUE)
plot(interval_avg_steps$interval, interval_avg_steps$steps,
        type = "l", 
        xlab = "5-Minute Interval",
        ylab = "Average Steps Across All Days",
        main = "Average Steps by 5-Minute Interval",
        col = "lightblue", lwd = 2)
```
which 5 minute interval contain the highest number of steps
```{r}
print(paste('The maxium number of total steps is in interval:',
        interval_avg_steps[max(interval_avg_steps$steps),]$interval))
```

## The Third Set Of Questions

To deal with the missing values in the data, means of the intervals were used.
```{r}
table(rowSums(is.na(df))) # one means the row contains NA values
third_df <- df # Again this was done to avoid messing with original data

mean_values <- aggregate(third_df$steps~third_df$interval, FUN = mean, na.rm = T)
colnames(mean_values) <- c('interval','means')

clean_data <- merge(third_df, mean_values, by.x= 'interval', by.y = 'interval')
clean_data$steps <- ifelse(is.na(clean_data$steps), 
        clean_data$means, clean_data$steps)
clean_data$means<- NULL
clean_data$steps <- round(clean_data$steps, 2)
```

Calculating the total amount of steps per day
```{r}
total_steps_clean <- tapply(clean_data$steps, clean_data$date, sum)
total_steps_clean <- as.data.frame(total_steps_clean)
```

Compression of the plots before and after
```{r}
par(mfcol = c(1,2), mar = c(4,4,3,2))
hist(total_steps_clean$total_steps_clean, col = color, 
        xlab = 'total steps per day', main = 'Total steps taken each day 
        (clean data)')
hist(sum_data$step_sum,col = color, xlab = 'Sum of the steps',
        main = 'Total Steps per Day (Data With NA)')
```

difference between the mean and median values after the data was cleaned
```{r}
clean_mean <- mean(total_steps_clean$total_steps_clean)
clean_median <- median(total_steps_clean$total_steps_clean)

print(paste('The mean value after cleaning the data is:' ,
        round(clean_mean,2), 'before it was:', round(mean_value,2)))
print(paste('The median value after cleaning the data is:' ,
        round(clean_median,2), 'before it was:',round(median_value,2)))
```

After cleaning the data by imputing missing values, the mean number of steps per 
day increased from 9354.23 to 10766.18, and the median number of steps per 
day rose from 10395 to 10766.13. This suggests that imputing the missing 
values with the interval mean resulted in a slight upward shift in the overall 
daily step counts.


## Last Set Of Questions

Creating the weekends columns, this column is used to compare the total steps
in the weekends and the weekdays.
```{r}
library(lubridate)
clean_data$weekday <- wday(clean_data$date, label = T)

clean_data$weekday <- ifelse(clean_data$weekday == 'Sat' 
        | clean_data$weekday == 'Sun', 'weekends', 'weekdays')
```

Creating the plots
please be aware that this plot differs from the instructor plot, 
due to difference in the data  as mentioned by the instructor.
```{r}
clean_data$weekday <- as.factor(clean_data$weekday)
par(mfrow = c(2,1), mar = c(4,4,4,4))
with(subset(clean_data, weekday == 'weekdays'), plot( interval, log(steps),
        col = 'lightblue', type = 'l',
        xlab = 'interval',
        ylab = 'Total Steps',
        main = 'Total Steps in the weekdays'))
with(subset(clean_data, weekday == 'weekends'), plot( interval, log(steps),
        col = 'lightblue', type = 'l',
        xlab = 'interval',
        ylab = 'Total Steps',
        main = 'Total Steps in the weekends '))
```